Detecting Hate Speech with GPT-3

### 2022-03-16 outputs/bmc_template/bmc_article.tex
Final read of paper

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   corref={aff1},                       % id of corresponding address, if any
   noteref={n1},                        % id's of article notes, if any
   email={rohan.alexander@utoronto.ca}   % email address
]{\inits{KC}\fnm{Ke-Li} \snm{Chiu}}
\author[
   addressref={aff1,aff2},
   email={rohan.alexander@utoronto.ca}
]{\inits{RA}\fnm{Rohan} \snm{Alexander}}

@@

\begin{abstract} % abstract
Sophisticated language models such as OpenAI's GPT-3 can generate hateful text that targets marginalized groups. Given this capacity, we are interested in whether large language models can be used to identify hate speech and classify text as sexist or racist? We use GPT-3 to identify sexist and racist text passages with zero-, one-, and few-shot learning. We find that with zero- and one-shot learning, GPT-3 can identify sexist or racist text with an accuracy between 48 per cent and 69 per cent. With few-shot learning and an instruction included in the prompt, the model's accuracy can be as high as 78 per cent. We conclude that large language models have a role to play in hate speech detection, and that with further development language models could be used to counter hate speech and even self-police.
\end{abstract}

@@

Natural language processing (NLP) models focus on using normal words, often written text, as their data. For instance, one might have content from a large number of books and want to group them into themes. Sophisticated NLP models are being increasingly embedded in society. For instance, Google Search uses an NLP model called BERT to better understand what is meant by a word given its context. Some sophisticated natural language processing (NLP) models, such as OpenAI's GPT-3, can additionally produce text as an output.

The text produced by sophisticated NLP models can be hateful. In particular, there have been many examples of text being generated that targets marginalized groups based on their sex, race, sexual orientation, and other characteristics. For instance, `Tay' was a chatbot that was released by Microsoft in 2016 on Twitter. Within hours of being released, some of its tweets were sexist. Large language models are trained on enormous datasets from various sources. This means that untruthful statements, human biases, and abusive language are inevitably included. Further, even when the models do not possess intent, they risk producing synthetic texts that are offensive or discriminatory and thus cause unpleasant, or even triggering, interaction experiences \cite{bender2021dangers}.

Often the datasets that underpin these models consist of, essentially, the whole public internet. This source raises concerns around three issues: exclusion, over-generalization, and exposure \cite{hovy2016social}. Exclusion happens due to the demographic bias in the dataset. In the case of language models that are trained on US and UK English scraped from the Internet, datasets may be disproportionately white, male, and young. Therefore, it is not surprising to see white supremacist, misogynistic, and ageist content being over-represented in training datasets \cite{bender2021dangers}. Over-generalization stems from the assumption that what we see in the dataset represents what occurs in the real world. Words such as `always', `never', `everybody', or `nobody' are frequently used for rhetorical purpose instead of their literal meanings. However, language models do not always recognize this and make inference based on generalized statements using these words. For instance, hate speech commonly uses generalized language for targeting a group such as `all' and `every', and a model trained on these statements may generate similarly generalized and harmful statements. Finally, exposure refers to the relative attention, and hence considerations of importance, given to something. In the context of NLP this may be reflected in the emphasis on English-language created under particular circumstances, rather than another languages or circumstances that may be more prevalent.

While these, and other, issues give us pause, the dual-use problem, which is that the same technology can be applied to both good and bad uses, provides motivation. For instance, while stylometric analysis can reveal the identity of political dissenters, it can also solve the unknown authorship of historic text \cite{hovy2016social}. In this paper we are interested in whether given large language models can produce harmful language, can they also identify, or learn to identify, harmful language?

Even though large NLP models do not have a real understanding of language, the vocabularies and the construction patterns of harmful languages can be thought of as known to them. We show that this knowledge can be used to identify abusive language and even hate speech. In particular we consider 120 different extracts that have been categorized as `racist', `sexist', or `neither' in the single-category settings and 240 different extracts in the mixed-category settings. We ask GPT-3 to classify these based on zero-, one-, and few-shot learning, with and without instruction. We find that the model performs best with few-shot learning when an instruction is included. In that setting the model can accurately classify around 78 per cent of the extracts. If language models can be used to identify abusive language, then not only is there potential for them to counter the production of abusive language by humans, but they could also potentially self-police.

@@

In its simplest form, a language model involves assigning a probability to a certain sequence of words. For instance, the sequence `the cat in the hat' is probably more likely than `the cat in the computer'. We typically talk of tokens, or collections of characters, rather than words, and a sequence of tokens constitutes different linguistic units --- words, sentences, and even documents \cite{bengio2003neural}. Language models predict the next token based on inputs. If we consider each token in a vocabulary as a dimension, then the dimensionality of language quickly becomes large \cite{rosenfeld2000two}. Over time a variety of statistical language models have been created to nonetheless enable prediction. The n-gram is one of the earliest language models. It works by considering the co-occurrence of tokens in a sequence. For instance, given the four-word sequence, `the cat in the', it is more likely that the fifth word is `hat' rather than `computer'. In the early 2000s, language models based on neural networks were developed, for instance \cite{bengio2003neural}. These were then built on by word embeddings language models in the 2010s in which the distance between tokens represents how related those tokens are, for instance \cite{turian2010word}. In 2017, \cite{vaswani2017attention} introduced the Transformer, which marked a new era for language models. The Transformer is a network architecture for neural networks that can be trained more quickly than many other approaches \cite{vaswani2017attention}. Now most representative pre-trained language models, such as Google's model: Bidirectional Encoder Representations from Transformers (BERT) \cite{devlin2018bert}, as well as OpenAI's: Generative Pre-trained Transformer (GPT)-2 \cite{radford2019language}, and GPT-3 \cite{brown2020language}, are built on this architecture. These models are widely used, for instance BERT is used by Google search.

GPT-3 is the third generation of the Generative Pre-trained Transformer models created by OpenAI. OpenAI is a private company in California that develops artificial intelligence models. Until January 11, when \cite{fedus2021switch} announced a Transformer model with a trillion parameters, GPT-3 was the largest, publicly-known, Transformer language model. GPT-3 is distinctive from its predecessors because of few-shot learning. This means that GPT-3 can `learn' to perform a new task based on only a few examples, expressed in natural language, instead of a fine-tuning process that can require a large amount of data. GPT-3 has led to unexpected NLP applications, such as computational code generation given natural language prompts.

@@

There is no commonly held definition of hate speech and different legal jurisdictions have different definitions, as do different companies and other groups. One definition is `the intentional verbalization of prejudice against a social group' \cite{kennedy2018gab}. Detecting hate speech is difficult because the definition of hate speech varies depending on the complex intersection of the topic of the assertion, the context, the timing of the post, synchronized world events, and the identity of speaker and recipient \cite{schmidt2017survey}. Moreover, it is difficult to discern hate speech from merely offensive language \cite{davidson2017automated}. Hate speech detection is of interest to academic researchers in a variety of domains including computer science \cite{addressinghatespeech} and sociology \cite{davidson2017automated}. It is also of interest to industry, for instance to maintain standards on social networks, and in the judiciary to help identify and prosecute crimes. Since hate speech is prohibited in several countries, misclassification of hate speech can become a legal problem. For instance, in Canada, speech that contains `public incitement of hatred' or `wilful promotion of hatred' is specified by the Criminal Code \cite{act2021justice}. Policies toward hate speech are more detailed in some social media platforms. For instance, the Twitter Hateful Conduct Policy states:

@@

There has been a large amount of research focused on detecting hate speech. And as part of this various hate speech datasets have been created and examined. For instance, \cite{waseem2016hateful} create a dataset that captures hate speech in the form of racist and sexist language that includes domain expert annotation. The use Twitter data, and annotate 16,914 tweets: 3,383 as sexist, 1,972 as racist, and 11,559 as neither. They had a high degree of annotator agreement, and most of the disagreements were to do with sexism, and often explained by an annotator lacking apparent context. And, \cite{davidson2017automated} trains a classifier to distinguish between hate speech and offensive language. To define hate speech, they use online `hate speech lexicon containing words and phrases identified by internet users as hate speech'. It is important to note that even these datasets have bias. For instance, \cite{davidson2019racial} found racial bias in five different sets of Twitter data annotated for hate speech and abusive language. They found that tweets written in African American English are more likely to be labeled as abusive.

@@

We use the ETHOS dataset created by \cite{mollas2020ethos}. ETHOS is based on comments found in YouTube and Reddit. The ETHOS YouTube data is collected through Hatebusters \cite{anagnostou2018hatebusters}. Hatebusters is a platform that collects comments from YouTube and assigns a `hate' score to them using a support vector machine. That hate-score is only used to decide whether to consider the comment further or not. The Reddit data is collected from the Public Reddit Data Repository \cite{baumgartner2020pushshift}. The classification is done by contributors to a crowd-sourcing platform. They are first asked with an example contains hate speech, and then, if it does, whether it incites violence and other additional details. The dataset has two variants: binary and multi-label. In the binary dataset comments are classified as hate or non-hate based. In the multi-label variant, the comments are evaluated on measures that include violence, gender, race, ability, religion, and sexual orientation. The examples provided in this paper are from the ETHOS dataset and hence contain typos, misspelling, and offensive content.

We use all of the 998 statements in the ETHOS dataset that have a binary classification of hate speech or not hate speech. Of these, the 433 statements that contain hate speech additionally have multiple labels that classify the content, for instance, does the comment have to do with violence, gender, race, nationality, disability, etc. We considered all of the 136 statements that contain race-based hate speech, and we focus on the 76 whose race-based score is at least 0.5. Similarly, we considered the all of 174 statements that contain gender-based hate speech, and again focused on the 84 whose gender-based score is at least 0.5. To create a balanced dataset we randomly draw 120 of these statements, using stratified sampling, to result in 60 sexist statements and 60 racist statements. For comparison purposes, we randomly draw 120 statements that are classified as not hate speech. The balanced sample enables us to more easily understand our results.

@@

With one-shot learning, similarly to the zero-shot learning case, the model is asked to identify one passage of text in one category at a time. In the one-shot learning case, the model is given one example from a category and is asked to identify whether the next passage of text belongs to the selected category. Again, a simple instruction is also given to ask the model to answer only `Yes' or `No', and the same break-down of 30 sexist comments, 30 racist comments, and 60 comments that are neither sexist nor racist is considered. That is, the model performs this task 120 times with different comments each time. An example of one-shot learning for recognizing a comment that is labeled as sexist is:

@@

Under the few-shot setting, three examples are provided to the model. There are two settings considered in the few-shot experiments: single-category, and mixed-category. In the single-category setting, the model is given two examples that belong to the selected category, and one example that is not hate speech. The answer is expected to be binary. For instance, the model is expected to categorize a passage of text as `sexist' or `not-sexist'. An example of this is:

@@

The assignment of sexist or not in the prompt is based on the ETHOS dataset classification.

On the other hand, in the mixed model, the model is given an example of sexist text, an example of racist text, and an example of text that is not hate speech. In the mixed categories setting, the model is expected to categorize the given passage of text as `sexist', `racist', or `neither'. There are ten sets of examples; each set of examples is used for the model to classify 60 racist comments, 60 sexist comments, and 120 comments that are in neither category This results in 2,400 responses in total. The ten sets of examples are available in Table \ref{tab:appendixa} in the Appendix. An example of this is:

@@

\section{Result}\label{result}

@@

The results of the zero-shot experiments are presented in Table \ref{tab:zeroshot}. The model has 35 matches and 25 mismatches in the sexist category, and 23 matches and 37 mismatches in the racist category. The model performs better when identifying sexist comments compared with identifying racist comments. However, the overall ratio of matches and mismatches is 58:62. In other words, the accuracy in identifying hate speech in the zero-shot setting is 48 per cent.

@@

The results of the one-shot learning experiments are presented in Table \ref{tab:oneshot}. The model has 46 matches and 14 mismatches in the racist category, and 37 matches and 23 mismatches in the sexist category. In contrast with the result generated from zero-shot learning, the model performs slightly better in identifying racist comments compared with identifying sexist comments. The general performance in the one-shot setting is also better than in the zero-shot setting. The overall ratio of matches and mismatches is 83:37. In other words, the accuracy of identifying hate speech in the one-shot setting is 69 per cent.

@@

The results of the single-category, few-shot learning, experiments are presented in Table \ref{tab:fewshotsingle}. The model has 41 matches and 19 mismatches in the racist category, and 42 matches and 18 mismatches in the sexist category. Here, the model performs almost equally well at identifying racist comments compared with identifying sexist comments. The general performance in the few-shot learning setting is similar to the performance in the one-shot learning setting. The overall ratio of matches and mismatches is 83:37, or around 69 per cent, however the composition is different, compared with the one-shot learning setting.

@@

The results of the mixed-category few-shot experiments are presented in Table \ref{tab:fewshotmixed}. Among the ten sets of examples, Example Set 2 yields the best performance, as the model has the highest number of matches, 180, and the lowest number of mismatches, 60. In other words, the accuracy under the mixed category few-shot setting with Example Set 2 is 75 per cent. The example set that yields the worst results is Example Set 9, in which there are 137 matches and 103 mismatches. The accuracy rate with Example Set 9 is 57 per cent. The difference between Example Sets 2 and 9 suggests that, although the models are provided with same number of examples, the content of the examples also affects how the model makes inferences.

@@

@@

To reduce the chance of the model generating answers that are out of scope, a brief instruction is added to the prompt, specifying that the answers be: `sexist', `racist', or `neither'. The addition of an instruction successfully restricts the generated answers within the specified terms. The unique generated answers are: `racist', `neither', and `sexist'.

The results of the mixed-category, few-shot learning, with instruction, experiments are presented in Table \ref{tab:fewshotmixedinstruct}. With the addition of an instruction in the prompt, the example set that yields the best result is Example Set 1 instead of Example Set 2. The addition of the instruction in the prompt increases the highest number of matches from 180 to 187; the highest accuracy rate is increased from 75 per cent to 78 per cent. In almost all cases the accuracy of the model increases (Figure \ref{fig:comparison}). Across all the example sets, there are 1,785, or 74 per cent, matches and 615, or 26 per cent, mismatches.

@@

In the zero-shot learning setting where the model is given no examples, its accuracy rate for identifying sexist and racist text is 48 per cent. In both the one-shot learning setting and the single-category few-shot learning setting, the accuracy increases to 69 per cent. It is likely that the model is not suitable for use in hate speech detection in either the zero-shot learning or one-shot learning settings, as the accuracy rates are between 48.3 per cent and 69 per cent. \cite{davidson2017automated}, using a different model and approach, similarly find `that almost 40 per cent of hate speech is misclassified'. And when \cite{schick2021selfdiagnosis} use GPT-2 they find a similar ability to recognise sexually explicit content, however using an alternative model - Google's T5 \cite{raffel2020exploring} - they find better results.

In the mixed-category few-shot setting, different example sets yield different accuracy rates. Without an instruction included in the prompt, the accuracy rates range from 57 per cent (Example Set 9) to 75 per cent (Example Set 2). Adding an instruction improves the performance, and the highest accuracy rate increases from 75 per cent (Example Set 2) to 78 per cent (Example Set 1). Overall accuracy rates also increase in almost every example set after adding an instruction in the prompt. The large difference in accuracy rates between the example sets suggests that the content of the examples may matter. It may be that deliberately curated examples would enable better accuracy.

Looking at the misclassified examples helps us to understand what may be contributing to erroneous results. The following comment has a true label of `racist', but the model classifies it as `sexist'. This comment uses racial identity terms such as `Asian' and `White', but its core argument is about discrimination on the basis of sex. The model did not have the necessary sensitivity to categorize the comment as sexist instead of racist.

>>>>

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   corref={aff1},                       % id of corresponding address, if any
   noteref={n1},                        % id's of article notes, if any
   email={rohan.alexander@utoronto.ca}   % email address
]{\inits{KC}\fnm{Ke-Li} \snm{Chiu}}
\author[
   addressref={aff2},
   email={rohan.alexander@utoronto.ca}
]{\inits{AC}\fnm{Annie} \snm{Collins}}
\author[
   addressref={aff1,aff2},
   email={rohan.alexander@utoronto.ca}
]{\inits{RA}\fnm{Rohan} \snm{Alexander}}

@@

\begin{abstract} % abstract
Sophisticated language models such as OpenAI's GPT-3 can generate hateful text that targets marginalized groups. Given this capacity, we are interested in whether large language models can be used to identify hate speech and classify text as sexist or racist. We use GPT-3 to identify sexist and racist text passages with zero-, one-, and few-shot learning. We find that with zero- and one-shot learning, GPT-3 can identify sexist or racist text with an accuracy between 57 per cent and 68 per cent depending on the category of text and type of learning. With few-shot learning, the model's accuracy can be as high as 88 per cent. Large language models have a role to play in hate speech detection, and with further development could eventually be used to counter hate speech.
\end{abstract}

@@

Natural language processing (NLP) models use normal words, often written text, as their data. For instance, one might have content from many books and want to group them into themes. Sophisticated NLP models are being increasingly embedded in society. For instance, Google Search uses an NLP model called BERT to better understand what is meant by a word given its context. Some sophisticated NLP models, such as OpenAI's Generative Pre-trained Transformer 3 (GPT-3), can additionally produce text as an output.

The text produced by sophisticated NLP models can be hateful. In particular, there have been many examples of text being generated that targets marginalized groups based on their sex, race, sexual orientation, and other characteristics. For instance, `Tay' was a chatbot that was released by Microsoft in 2016 on Twitter. Within hours of being released, some of its tweets were sexist. Large language models are trained on enormous datasets from various sources. This means they usually contain untruthful statements, human biases, and abusive language. Further, even when the models do not possess intent, they risk producing text that is offensive or discriminatory, and thus cause unpleasant, or even triggering, interactions \cite{bender2021dangers}.

Often the datasets that underpin these models consist of, essentially, the whole public internet. This source raises concerns around three issues: exclusion, over-generalization, and exposure \cite{hovy2016social}. Exclusion happens due to the demographic bias in the dataset. In the case of language models that are trained on US and UK English scraped from the Internet, datasets may be disproportionately white, male, and young. Therefore, it is not surprising to see white supremacist, misogynistic, and ageist content being over-represented in training datasets \cite{bender2021dangers}. Over-generalization stems from the assumption that what we see in the dataset represents what occurs in the real world. Words such as `always', `never', `everybody', or `nobody' are frequently used for rhetorical purpose instead of their literal meanings. However, language models do not always recognize this and make inference based on generalized statements using these words. For instance, hate speech commonly uses generalized language for targeting a group such as `all' and `every', and a model trained on these statements may generate similarly generalized and harmful statements. Finally, exposure refers to the relative attention, and hence consideration of importance, given to something. In the context of NLP this may be reflected in the emphasis on English-language created under particular circumstances, rather than another language or circumstances that may be more prevalent.

While these issues, among others, give us pause, the dual-use problem, which is that the same technology can be applied for both good and bad uses, provides motivation. For instance, while stylometric analysis can reveal the identity of political dissenters, it can also solve the unknown authorship of historic text \cite{hovy2016social}. In this paper we are interested in whether large language models, given that they can produce harmful language, can also identify, or learn to identify, harmful language.

Even though large NLP models do not have a real understanding of language, the vocabularies and the construction patterns of harmful languages can be thought of as known to them. We show that this knowledge can be used to identify abusive language and even hate speech. We consider 120 different extracts that have been categorized as `racist', `sexist', or `neither' in single-category settings (zero-shot, one-shot, and few-shot) and 243 different extracts in mixed-category few-shot settings. We ask GPT-3 to classify these based on zero-, one-, and few-shot learning, with and without instruction. We find that the model performs best with mixed-category few-shot learning. In that setting the model can accurately classify around 86 per cent of the racist extracts and 88 per cent of sexist extracts. If language models can be used to identify abusive language, then not only is there potential for them to counter the production of abusive language by humans, but they could also potentially self-police.

@@

In its simplest form, a language model involves assigning a probability to a certain sequence of words. For instance, the sequence `the cat in the hat' is probably more likely than `the cat in the computer'. We typically talk of tokens, or collections of characters, rather than words, and a sequence of tokens constitutes different linguistic units --- words, sentences, and even documents \cite{bengio2003neural}. Language models predict the next token based on inputs. If we consider each token in a vocabulary as a dimension, then the dimensionality of language quickly becomes large \cite{rosenfeld2000two}. Over time a variety of statistical language models have been created to nonetheless enable prediction. The n-gram is one of the earliest language models. It works by considering the co-occurrence of tokens in a sequence. For instance, given the four-word sequence, `the cat in the', it is more likely that the fifth word is `hat' rather than `computer'. In the early 2000s, language models based on neural networks were developed, for instance \cite{bengio2003neural}. These were then built on by word embeddings language models in the 2010s in which the distance between tokens represents how related those tokens are, for instance \cite{turian2010word}. In 2017, \cite{vaswani2017attention} introduced the Transformer, which marked a new era for language models. The Transformer is a network architecture for neural networks that can be trained more quickly than many other approaches \cite{vaswani2017attention}. Now most representative pre-trained language models, such as Google's: Bidirectional Encoder Representations from Transformers (BERT) \cite{devlin2018bert}, as well as OpenAI's: Generative Pre-trained Transformer (GPT)-2 \cite{radford2019language}, and GPT-3 \cite{brown2020language}, are built on this architecture. These models are widely used, for instance BERT is used by Google search.

GPT-3 is the third generation of the Generative Pre-trained Transformer models created by OpenAI, a private company in California that develops artificial intelligence models. GPT-3 is an autoregressive NLP model that can perform a variety of tasks, including responding to questions, summarizing, and parsing text, translation, and classification. Interactions with the model involve inputting some text as a prompt and GPT-3 returning a text completion according to that prompt.

GPT-3 is one of the largest, publicly available, Transformer language models. GPT-3 is distinctive to its predecessors because of few-shot learning. This means that GPT-3 can `learn' to perform a new task based on only a few examples, expressed in natural language, instead of a fine-tuning process that can require a large amount of data. GPT-3 has led to unexpected NLP applications, such as computational code generation given natural language prompts.

@@

There is no commonly held definition of hate speech. Different legal jurisdictions have different definitions, as do different companies and other groups. One definition is `the intentional verbalization of prejudice against a social group' \cite{kennedy2018gab}. Detecting hate speech is difficult because the definition of hate speech varies depending on a complex intersection of the topic of the assertion, the context, the timing of the post, synchronized world events, and the identity of speaker and recipient \cite{schmidt2017survey}. Moreover, it is difficult to distinguish hate speech from offensive language \cite{davidson2017automated}. Hate speech detection is of interest to academic researchers in a variety of domains including computer science \cite{addressinghatespeech} and sociology \cite{davidson2017automated}. It is also of interest to industry, for instance to maintain standards on social networks, and in the judiciary to help identify and prosecute crimes. Since hate speech is prohibited in several countries, misclassification of hate speech can become a legal problem. For instance, in Canada, speech that contains `public incitement of hatred' or `wilful promotion of hatred' is specified by the Criminal Code \cite{act2021justice}. Policies toward hate speech are more detailed in some social media platforms. For instance, the Twitter Hateful Conduct Policy states:

@@

There has been a large amount of research focused on detecting hate speech. As part of this process, various hate speech datasets have been created and examined. For instance, \cite{waseem2016hateful} create a dataset that captures hate speech in the form of racist and sexist language that includes domain expert annotation. They use Twitter data, and annotate 16,914 tweets: 3,383 as sexist, 1,972 as racist, and 11,559 as neither. They had a high degree of annotator agreement. Most of the disagreements were to do with sexism, and often explained by an annotator lacking apparent context. \cite{davidson2017automated} trains a classifier to distinguish between hate speech and offensive language. To define hate speech, they use an online `hate speech lexicon containing words and phrases identified by internet users as hate speech'. Even these datasets have bias. For instance, \cite{davidson2019racial} found racial bias in five different sets of Twitter data annotated for hate speech and abusive language. They found that tweets written in African American English are more likely to be labeled as abusive.

@@

We use the ETHOS dataset created by \cite{mollas2020ethos}. ETHOS is based on comments found in YouTube and Reddit. The ETHOS YouTube data is collected through Hatebusters \cite{anagnostou2018hatebusters}. Hatebusters is a platform that collects comments from YouTube and assigns a `hate' score to them using a support vector machine. That hate-score is only used to decide whether to consider the comment further or not. The Reddit data is collected from the Public Reddit Data Repository \cite{baumgartner2020pushshift}. The classification is done by contributors to a crowd-sourcing platform. They are first asked whether an example contains hate speech, and then, if it does, whether it incites violence and other additional details. The dataset has two variants: binary and multi-label. In the binary dataset comments are classified as hate or non-hate based. In the multi-label variant comments are evaluated on measures that include violence, gender, race, ability, religion, and sexual orientation. The dataset that we use is from the ETHOS dataset and hence contain typos, misspelling, and offensive content.

We use all of the 998 statements in the ETHOS dataset that have a binary classification of hate speech or not hate speech. Of these, the 433 statements that contain hate speech additionally have labels that classify the content. For instance, does the comment have to do with violence, gender, race, nationality, disability, etc. We considered all of the 136 statements that contain race-based hate speech, and we focus on the 76 whose race-based score is at least 0.5, meaning that at least 50 per cent of annotators agreed. Similarly, we considered all of the 174 statements that contain gender-based hate speech, and again focused on the 84 whose gender-based score is at least 0.5. To create a balanced dataset, we select 120 of these statements to result in 60 sexist statements and 60 racist statements. For comparison purposes, we randomly draw 120 statements that are classified as not hate speech. The balanced sample enables us to more easily understand our results.

For the purposes of one-shot and single-category few-shot learning, the dataset was split in half such that 30 racist comments, 30 sexist comments, and 60 neutral comments were used as inputs and the remainder were used as examples. This division was kept consistent for zero-shot learning to allow for the same set of 120 comments to be classified by all three approaches. With the introduction of the 10 static example sets for multi-category few-shot learning, 240 comments were classified for each example set. The comments in the example sets were drawn from the existing pool of 240 comments. For each example set, the comment from each category being used in the example set was dropped from the input dataset and an additional comment from each original dataset (one `racist', one `sexist', and one `neither') was added in as a replacement. This replacement was kept consistent each time, for a total of 243 unique comments classified in the few-shot mixed-category setting.

@@

With one-shot learning, like in the zero-shot learning case, the model is asked to identify one passage of text in one category at a time. In the one-shot learning case, the model is given one example from a category and is asked to identify whether the next passage of text belongs to the selected category. Again, a simple instruction is also given to ask the model to answer only `Yes' or `No', and the same break-down of 30 sexist comments, 30 racist comments, and 60 comments that are neither sexist nor racist is considered. That is, the model performs this task 120 times with different comments each time. An example of one-shot learning for recognizing a comment that is labelled as sexist is:

@@

Under the few-shot setting, three examples are provided to the model. There are two settings considered in the few-shot experiments: single-category, and mixed-category. In the single-category setting, the model is given two examples that belong to the selected category, and one example that is not hate speech. The answer is expected to be binary. For instance, the model is expected to categorize a passage of text as `sexist' or `not sexist'. An example of this is:

@@

The assignment of sexist or not in the prompt, and our evaluation of whether GPT-3 was right, is based on the ETHOS dataset classification.

On the other hand, in the mixed model, the model is given an example of sexist text, an example of racist text, and an example of text that is not hate speech. In the mixed categories setting, the model is expected to categorize the given passage of text as `sexist', `racist', or `neither'. To simplify the classification procedure and allow for comparisons between different example inputs, we created ten sets of examples against which 240 comments were each classified. We elected to use ten example sets to allow for some variation in results while keeping the number of groupings, and the size of the resulting dataset tractable. The example sets each consist of one sexist, one racist, and one `neither' comment and were created using the comments at indexes one through ten from each dataset. Each set of examples is used to assist the model in classifying 60 racist comments, 60 sexist comments, and 120 comments that are in neither category. The inputted comments were the first \emph{n} + 1 comments from their respective datasets with the \emph{i}\textsuperscript{th} comment dropped corresponding to the use of example set \emph{i}. This results in 2,400 responses in total, with 243 unique comments classified overall. The ten sets of examples are shown in Appendix \ref{appendxa}. An example of this is:

@@

\section{Results}\label{results}

We assess GPT-3's performance in all settings using accuracy, precision, recall, and F1 score. Accuracy is the proportion of correctly classified comments (hate speech and non-hate speech) out of all comments classified. Precision is the proportion of hate speech comments correctly classified out of all comments classified as hate speech (both correctly and incorrectly). Recall is the proportion of hate speech comments correctly classified out of all hate speech comments in the dataset (both correctly and incorrectly classified). The F1 score is the harmonic mean of precision and recall. In the case of hate speech classification, it is better to have a model with high recall, meaning a model that can identify a relatively high proportion of the hate speech text within a dataset. But the F1 score can provide a more well-rounded metric for model performance and comparison.

@@

The overall results of the zero-shot experiments are presented in Table \ref{tab:zeroshot-summary}. Tables \ref{tab:zeroshot-racism}, \ref{tab:zeroshot-sexism}, and \ref{tab:zeroshot-hate}, in Appendix \ref{appendixbzeroshot} contain additional detail. The model has 34 matches (true positives and negatives) and 26 mismatches (false positives and negatives) in the sexist category, and 41 matches and 19 mismatches in the racist category. The model performs more accurately when identifying racist comments, with an accuracy of 68 per cent, compared with identifying sexist comments, with an accuracy of 57 per cent, however the F1 score for each type of hate speech is approximately the same. The overall ratio of matches and mismatches is 75:45. In other words, the accuracy in identifying hate speech in the zero-shot setting is 63 per cent. The model has an F1 score of 81 per cent in this setting, with a notably high recall of 87 per cent as well.

@@

The results of the one-shot learning experiments are presented in Table \ref{tab:oneshot-summary}. Tables \ref{tab:oneshot-racism}, \ref{tab:oneshot-sexism}, and \ref{tab:oneshot-hate}, in Appendix \ref{appendixboneshot} provide additional detail. The model has 37 matches and 23 mismatches in the racist category, and 36 matches and 24 mismatches in the sexist category. In contrast with the result generated from zero-shot learning, the model performs approximately the same in identifying racist comments and identifying sexist comments. The overall ratio of matches and mismatches is 73:47. In other words, the accuracy of identifying hate speech in the one-shot setting is 61 per cent. The general performance in the one-shot setting is worse than in the zero-shot setting, with lower accuracy, recall, precision, and F1 score overall.

@@

The results of the single-category, few-shot learning, experiments are presented in Table \ref{tab:fewshotsingle-summary}. Tables \ref{tab:fewshotsingle-racism}, \ref{tab:fewshotsingle-sexism}, \ref{tab:fewshotsingle-hate}, in Appendix \ref{appendixbfewshotsingle} provide additional detail. The model has 41 matches and 19 mismatches in the racist category, and 42 matches and 18 mismatches in the sexist category. Like the one-shot setting, the model performs almost equally well at identifying racist comments compared with identifying sexist comments. The general performance in the single-category few-shot learning setting is slightly more accurate than performance in other settings, with an accuracy of 69 per cent compared with 61 per cent in the one-shot setting and 63 per cent in the zero-shot setting. However, the model in this context has the lowest F1 score of the settings considered thus far at 61 per cent.

@@

The results of the mixed-category few-shot experiments are presented in Table \ref{tab:fewshotmixed-summary}. Tables \ref{tab:fewshotmixed-racism} and Table \ref{tab:fewshotmixed-sexism} in Appendix \ref{appendxbmuxedubsnorinstruction} provide additional detail. Among the ten sets of examples, Example Set 1 yields the best performance in terms of accuracy (91 per cent) and F1 score (87 per cent) for racist comments and Example Set 10 yields the best performance by the same metrics for sexist comments (accuracy of 89 per cent, F1 score of 85 per cent). The example set that yields the worst results in identifying racist text in terms of F1 score is Example Set 8, which has an F1 score of 70 per cent (and the lowest accuracy at 73 per cent) for this dataset. The example set that yields the worst results in identifying sexist text in terms of F1 score is Example Set 9, which has an F1 score of 69 per cent (and the lowest accuracy at 74 per cent) for this dataset. The differences between Example Sets 1, 8, 9, and 10 suggest that, although the models are provided with same number of examples, the content of the examples also affects how the model makes inferences. Overall, the mixed-category few-shot setting performs approximately the same at identifying sexist text and racist text. It is also has higher accuracy and F1 score than the zero-shot, one-shot, and single-category few-shot settings for both racist and sexist text.

@@

The results presented for each category of text include the classifications of comments that were labelled as `neither' and the category in question. For the purposes of our analysis, a classification was considered a true positive if the answer outputted by GPT-3 contained a category that matched the comment's label. For example, if a comment was labelled `sexist' and the comment was classified by the model as `sexist, racist', this was considered a true positive in the classification of sexist comments. If a comment was labelled `sexist' and the comment was classified by the model as `racist', `transphobic', `neither', etc. this was considered a false negative.

Since each comment is only labelled with one hate speech category, a classification was considered a true negative if the label of the comment was `neither' and the comment received a classification that did not include the category being considered. For example, if a comment was labelled `neither' and the model answered `racist', this is considered a true negative in the classification of sexist comments (the comment is not sexist, and the model did not classify it as sexist), but a false positive in the classification of racist comments (the comment is not racist, but the model classified it as racist).

@@

To reduce the chance of the model generating answers that are out of scope, a brief instruction is added to the prompt, specifying that the answers be: `sexist', `racist', or `neither'. The addition of an instruction successfully restricts the generated answers within the specified terms. The unique generated answers are: `racist', `sexist', and `neither'.

The results of the mixed-category few-shot learning, with instruction, experiments are presented in Tables \ref{tab:fewshotmixedinstruct-matrix} and \ref{tab:fewshotmixedinstructsummary}. And Tables \ref{tab:fewshotmixedinstruct-racism}, and \ref{tab:fewshotmixedinstruct-sexism}, in Appendix \ref{appendxbmuxedubstryctub} provide additional detail. With the addition of an instruction in the prompt, Example Set 10 remains the best performing example set in terms of accuracy (89 per cent) and F1 score (84 per cent) for sexist text, however Example Set 4 now performs best for racist text in terms of both accuracy (89 per cent) and F1 score (85 per cent). When considering the classification of racist and sexist speech overall, the models perform similarly with and without instruction when classifying racist text, but the model appears to perform slightly better at identifying sexist text when the instruction is omitted.

However, examining label-classification matches across all categories (`sexist', `racist', and `neither'), mixed-category few-shot learning almost always performs better with instruction than without instruction (Figure \ref{fig:comparison}). Across all example sets, the mean proportion of matching classifications (out of 240 comments) for mixed-category few-shot learning without instruction is 67 per cent. The average proportion of matching classifications rises to 75 per cent for learning with instruction.

@@

In the zero-shot learning setting where the model is given no examples, its accuracy rate for identifying sexist and racist text is 63 per cent with an F1 score of 81 per cent. In the one-shot learning setting the accuracy decreases to 61 per cent with an F1 score of 65 per cent. Accuracy increases to 69 per cent in the single-category few-shot learning setting, with an F1 score of 61 per cent. Of the three single-category approaches, zero-shot learning has a notably high recall value of 87 per cent.

It is likely that the model is not ideal for use in hate speech detection in the zero-shot learning, one-shot learning, or single-category few-shot learning settings, as the accuracy rates are between 60 per cent and 70 per cent. \cite{davidson2017automated}, using a different model and approach, similarly find `that almost 40 per cent of hate speech is misclassified'. And when \cite{schick2021selfdiagnosis} use GPT-2 they find a similar ability to recognize sexually explicit content, however using an alternative model - Google's T5 \cite{raffel2020exploring} - they find better results.

In the mixed-category few-shot setting, different example sets yield different accuracy rates for racist and sexist comments, with noticeable improvement over the single-category approaches. Mixed-category few-shot learning without instruction had noticeably better F1 scores for both racist and sexist comments than either zero-shot or one-shot learning. With instruction added, mixed-category few-shot learning performed similarly well for racist text identification, however the model performed relatively poorly in identifying sexist speech, with an F1 score of 63 per cent overall and a recall of 50 per cent meaning nearly half of the sexist comments were wrongly classified. Overall, it appears as though GPT-3 is most effective at identifying both racist and sexist comments in the mixed-category few-shot learning setting, without instruction.
